An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-06 main.tex
string length

Parikh images play a central role in many automata-based algorithms,
for instance and notably in today's string solvers, which often have
to process constraints that combine regular language membership with
word length. To decide whether a simple formula like
$x \in \mathcal{L}_1 \wedge y \in \mathcal{L}_2 \wedge |x| > |y|$,
with string variables~$x, y$ and regular
languages~$\mathcal{L}_1, \mathcal{L}_2$, is satisfiable, it is
necessary to reason about the sets of word lengths induced by
$\mathcal{L}_1, \mathcal{L}_2$, which is a special case of the Parikh
image.  This required combined reasoning about strings and string
length has long been identified as a major bottleneck in string
solvers
\cite{DBLP:conf/cav/AbdullaACHRRS15,length-aware-solver,approximate-parikh,more}.
Other string solvers make use of Parikh automata~\cite{X}, and thus
Parikh images in the general case, to handle operations including
\verb!str.substr!  and \verb!str.at!, which comes at an even higher
price in terms of computational complexity~\cite{ostrich-plus}.

@@

\subsection{Example~1: Reasoning about string length}

A useful example of such a simplifying morphism can express string
length, the problem that originally motivated our study of the Parikh
map. We let $\Monoid = \left(\mathbb{Z};+;0\right)$, and define the
homomorphism~$\Map$ by $\Map(a) = 1$ for all
characters~$a \in \Alphabet$. Then the length of a string
$s = s_1 \RepeatSum{\Concat} s_n$ is given by
$\Map(s) = \sum \Map(s_i) = 1 \RepeatSum{+} 1 = n$.

Intuitively, this produces a simpler problem than computing the Parikh image, as
we now ignore not just the order of characters but also their differences.
\iffalse
This
means that the branching automaton of \cref{fig:len-branch} becomes
equivalent to the non-branching automaton of \cref{fig:len-branch-free}.
Note that the elimination of the choice of paths also effectively eliminates the
possibility of disconnecting one of the self-loops. This, in turn, means that we
can eliminate all of the connectedness implications of the quantified formula,
which would in turn simplify it (and the quantifier elimination of it)
considerably.
\fi

\subsection{Example~2: Parikh Automata}\label{sec:parikh-automata}
\contents{\item Ostrich+!}

>>>>

Parikh images play a central role in many automata-based algorithms,
for instance and notably in today's string solvers, which often have
to process constraints that combine regular language membership with
word length. To decide whether a simple formula like
$x \in \Language_1 \wedge y \in \Language_2 \wedge |x| > |y|$,
with string variables~$x, y$ and regular
languages~$\Language_1, \Language_2$, is satisfiable, it is
necessary to reason about the sets of word lengths induced by
$\Language_1, \Language_2$, which is a special case of the Parikh
image.  This required combined reasoning about strings and string
length has long been identified as a major bottleneck in string
solvers
\cite{DBLP:conf/cav/AbdullaACHRRS15,length-aware-solver,approximate-parikh,more}.
Other string solvers make use of Parikh automata~\cite{X}, and thus
Parikh images in the general case, to handle operations including
\verb!str.substr!  and \verb!str.at!, which comes at an even higher
price in terms of computational complexity~\cite{ostrich-plus}.

@@

\subsection{Example~1: Reasoning about String Length}

A useful example of such a simplifying morphism can express string
length, the problem that originally motivated our study of the Parikh
map. This mapping is relevant when solving constraints that combine
language membership as well as word length, for instance the formula
\begin{equation}\label{eq:stringLength}
x \in \Language_1 \wedge y \in \Language_2 \wedge |x| > |y|
\end{equation}
mentioned in the introduction. To solve this formula, let
$\Monoid = \Naturals$, and define the homomorphism~$L$ by $L(a) = 1$
for all characters~$a \in \Alphabet$. The length of a string
$s = s_1 \RepeatSum{\Concat} s_n$ is given by
$L(s) = \sum L(s_i) = 1 \RepeatSum{+} 1 = n$, and to solve
\eqref{eq:stringLength} we can instead solve the equi-satisfiable
formula~$\alpha \in L(\Language_1) \wedge \beta \in L(\Language_2)
\wedge \alpha > \beta$. This paper proposes efficient native
procedures to reason about membership constraints
like~$\alpha \in L(\Language_1)$, avoiding the computation of
the complete image~$L(\Language_1)$.

Intuitively, this produces a simpler problem than computing the Parikh image, as
we now ignore not just the order of characters but also their differences.
This
means that the branching automaton of \cref{fig:len-branch} becomes
equivalent to the non-branching automaton of \cref{fig:len-branch-free}.
Note that the elimination of the choice of paths also effectively eliminates the
possibility of disconnecting one of the self-loops. This, in turn, means that we
can eliminate all of the connectedness implications of the quantified formula,
which would in turn simplify it (and the quantifier elimination of it)
considerably.

\subsection{Example~2: Generalised String Constraints with Integer
  Datatype}\label{sec:parikh-automata}
\contents{\item Ostrich+!}
